SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-05 back.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

Moreover, the on-line confidence mechanism used by most prefetchers is very
rudimentary. For example, SPP's internal confidence mechanism is based on taking 
the ratio C\textsubscript{d} = C\textsubscript{delta} / C\textsubscript{sig}.
This confidence was used to make the decision of whether to prefetch or not to
prefetch; and which level to prefetch.  While this approximation was shown to
work in the original implementation, we believe that a better form of
generalized on-line decision making was possible.  Hence, it was necessary to
build a robust and adaptable learning mechanism to accept / reject the
prefetch suggestions; and to decide the fill level (L2 Cache vs Last Level
Cache).

To that effect, we introduce an independent on-line perceptron based
filtering mechanism.

@@

%ETG: this is what I had added about perceptrons, we're trying to 
%figure out what to leave and what to get rid of ...  
\subsection{Perceptron-based Prediction}

Perceptrons in microarchitecture were first introduced by Jimenez and Lin~\cite{Perc_Branch} 
for branch prediction. In this work, a perceptron weights vector is selected by a hash of 
the branch address and its dot product with an input vector of the previous branch history 
outcomes. If the product is at least 0, the branch is predicted as taken, otherwise it is 
predicted not taken. The weights are updated with the perceptron update rule: if the prediction 
is incorrect, or the dot product does not exceed some threshold magnitude, then the weights 
are incremented or decremented based on their correlation to the corresponding history bits. 
Perceptron learning yields to superior accuracy, which results in increased opportunity for 
optimizations.

\begin{figure}[ht]
\includegraphics[width=\columnwidth]{Perc_Model}
\caption{The Basic Perceptron Model}
\label{Fig:Perc_Model}
\end{figure}

\subsubsection{Perceptron Learning}

In a perceptron-based mechanism, a uniform perceptron update principle is followed. As
shown in the figure~\ref{Fig:Perc_Model} Perceptrons 
are a simple model of neurons in neural networks \cite{perceptrons1,perceptrons2} modeled by 
vectors of signed weights learned through online training. The output of a perceptron is the 
dot product of the weights and a vector of inputs. In this work, we make use of the perceptron 
learning algorithm. There are two components to the abstract perceptron learning algorithm:

\begin{enumerate}
\item To predict true or false, a vector of signed weights is chosen according to some criteria. 
The dot product of the weights and input vectors, called $y_{\mbox{\small out}}$, is computed. 
If $y_{\mbox{\small out}}$ exceeds some threshold, the prediction is true, otherwise it is false.

\item To update the weights after the true outcome of the predicted event is
known, we first consider the value of $y_{\mbox{\small out}}$. If the
prediction was correct and $|y_{\mbox{\small out}}|$ exceeds a threshold
$\theta$, then the weights remain unchanged. Otherwise, the inputs are used to
update the corresponding weights.  If there is positive correlation between
the input and the outcome of the event, the corresponding weight is
incremented; otherwise, it is decremented. Over time, the weights are
proportional to the probability that the outcome of the event is true in the
context of the input and the criterion for choosing that weight. The weights
saturate at maximum and minimum values so they will fit in a fixed bit width.
\end{enumerate}

>>>>

Moreover, the on-line confidence mechanism used by most prefetchers is very
rudimentary. For example, SPP's internal confidence mechanism is based on
taking the ratio C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}.  This confidence was used to make the decision of
whether to prefetch or not to prefetch; and which level to prefetch.  While
this approximation was shown to work in the original implementation, we
believe that a better form of generalized on-line decision making was
possible.  Hence, it was necessary to build a robust and adaptable learning
mechanism to accept / reject the prefetch suggestions; and to decide the fill
level (L2 Cache vs Last Level Cache).  Thus, we introduce an independent
on-line perceptron based filtering mechanism.

@@

%ETG: this is what I had added about perceptrons, we're trying to 
%figure out what to leave and what to get rid of ...  
% djimenez: I ended up cutting everything you wrote as well as the figure. I
% wrote something similar above, and the perceptron model described here isn't
% the hashed perceptron that we're using but rather the dot-product model from
% the original paper.

%\subsection{Perceptron-based Prediction}
%
%Perceptrons in microarchitecture were first introduced by Jimenez and Lin~\cite{PerceptronPredictor} 
%for branch prediction. In this work, a perceptron weights vector is selected by a hash of 
%the branch address and its dot product with an input vector of the previous branch history 
%outcomes. If the product is at least 0, the branch is predicted as taken, otherwise it is 
%predicted not taken. The weights are updated with the perceptron update rule: if the prediction 
%is incorrect, or the dot product does not exceed some threshold magnitude, then the weights 
%are incremented or decremented based on their correlation to the corresponding history bits. 
%Perceptron learning yields to superior accuracy, which results in increased opportunity for 
%optimizations.

%\begin{figure}[ht]
%\includegraphics[width=\columnwidth]{Perc_Model}
%\caption{The Basic Perceptron Model}
%\label{Fig:Perc_Model}
%\end{figure}

%\subsubsection{Perceptron Learning}
%
%In a perceptron-based mechanism, a uniform perceptron update principle is
%followed. As shown in the figure~\ref{Fig:Perc_Model} Perceptrons are a simple
%model of neurons in neural networks \cite{perceptrons1,perceptrons2} modeled
%by vectors of signed weights learned through online training. The output of a
%perceptron is the dot product of the weights and a vector of inputs. In this
%work, we make use of the perceptron learning algorithm. There are two
%components to the abstract perceptron learning algorithm:
%
%\begin{enumerate}
%\item To predict true or false, a vector of signed weights is chosen according to some criteria. 
%The dot product of the weights and input vectors, called $y_{\mbox{\small out}}$, is computed. 
%If $y_{\mbox{\small out}}$ exceeds some threshold, the prediction is true, otherwise it is false.
%
%\item To update the weights after the true outcome of the predicted event is
%known, we first consider the value of $y_{\mbox{\small out}}$. If the
%prediction was correct and $|y_{\mbox{\small out}}|$ exceeds a threshold
%$\theta$, then the weights remain unchanged. Otherwise, the inputs are used to
%update the corresponding weights.  If there is positive correlation between
%the input and the outcome of the event, the corresponding weight is
%incremented; otherwise, it is decremented. Over time, the weights are
%proportional to the probability that the outcome of the event is true in the
%context of the input and the criterion for choosing that weight. The weights
%saturate at maximum and minimum values so they will fit in a fixed bit width.
%\end{enumerate}

### 2018-12-05 conc.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

\section{Conclusion}
\label{Conclusion}
In this paper, we introduced perceptron-based learning for data prefetching.
The perceptron filter acts independently as a check on the underlying prefetch
engine.  PPF improves performance of the base prefetcher used in isolation, 
by 11.9\%.  We believe that this is a robust and adaptable technique 
that can be used to enhance any existing and upcoming prefetchers.

>>>>

\section{Conclusion}
\label{Conclusion}
In this paper, we introduce perceptron-based learning for prefetching.  The
perceptron filter acts independently as a check on the underlying prefetch
engine.  PPF improves performance over the base prefetcher by 11.9\%.  We
believe that this is a robust and adaptable technique that can be used to
enhance any existing and upcoming prefetchers.

### 2018-12-05 design_arch.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{PPF_Hierarchy.pdf}
    \label{fig:PPF_Hierarchy}
  \caption{PPF Architecture in the Memory Hierarchy}
  \end{center}
\end{figure}

As mentioned before, ultimately it might be beneficial to allow a
prefetcher to speculate as deeply as possible. Often, some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below a point, after which the performance degrades, due to the
increase of inaccurate prefetches.  However, in order to allow deep
speculation in the prefetcher, inaccurate prefetches must be
filtered out.  We propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones. The Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, as illustrated
in figure ~\ref{fig:PPF_Hierarchy}, to prevent
not-useful prefetches from polluting the higher levels of the memory hierarchy.

As discussed in Section~\ref{sec:Background-Perceptron}, perceptron learning
is a light-weight mechanism to pull together disparate forms of information
and synthesize a decision from them.  Our work considers a number of
features corresponding to a prefetch, such as speculation
depth, page address and offset, and uses this information as the inputs to our 
perceptron-based filter, in order to predict the usefulness of a prefetch.
In the remainder of this section, we discuss our proposed technique -- PPF.
It enhances a base prefetcher, by filtering out not-useful prefetches. 
PPF is a generalized prefetch filtering mechanism that may be adapted 
to any prefetcher with appropriate modifications.

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as the
steps followed to filter out not-useful prefetches.
The filter is organized as a set of tables, where each entry in the tables holds 
a \textit{weight}. For a configuration of PPF using $N$ number of features, 
$N$ different tables of weights are needed. Each feature is used to index 
a distinct table. The number of entries of each table varies according
to their corresponding feature, hence, different
number of bits are needed to index different tables.
Each weight is a 5-bit saturating counter ranging from -16 to +15. 
We found that having 5-bit weights provide a good trade-off 
between accuracy and area. A detail explanation of the storage overhead of
PPF can be found in Section~\ref{sec:Method-Overheads}.

@@

\noindent{\textbf{1. Inference}}\newline

%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.

The base prefetcher is triggered on every demand access to the L2 Cache.  
As this point, it has the opportunity to trigger a prefetch  -- if it 
decides to do so, it will also need to decide how many cache lines
to prefetch.  These cache lines can be either placed in the L2 Cache or
the L3 Cache, according to the confidence of the prefetching mechanism.
Once the base prefetcher is triggered, the suggested
prefetch candidates are fed to the perceptron filter, 
in order to determine the usefulness of these prefetches.  
The filter makes the decision to issue the prefetch
suggestions from the base prefetcher.
In order to make such decision, each feature corresponding to a 
suggested prefetch is used to index a table, 
all the corresponding weights are summed. 
The sum denotes the confidence value for the suggested prefetch, 
and is thresholded against two different magnitudes:  
PERC\_THRESHOLD\_HI and PERC\_ THRESHOLD\_ LO.

Prefetches which sum is above PERC\_THRESHOLD\_HI
are placed into the L2 cache, the higher confidence value
hints the prefetch would be useful and should be prioritized.
A prefetch for which its corresponding features result 
in a confidence value between  
PERC\_THRESHOLD\_LO and PERC\_THRESHOLD\_HI is allocated in the larger 
last-level cache, as the filter is moderately confident of the future reuse of the 
cache line, but not enough to possible pollute a significantly smaller L2. 
Suggested prefetches for which the features lead to a confidence value lower than 
PERC\_THRESHOLD\_LO are not prefetched, as the low confidence value 
represents that the perceptron has learn that similar set of features
are associated with non-useful prefetches. \newline

\noindent{\textbf{2. Recording}}\newline

As shown in step 2 of figure ~\ref{fig:PPF_Datapath}
the prefetches that make it through the inference stage are recorded in the 
``Prefetch Table''. The prefetch table is a 1,024-entry, direct mapped 
structure that contains all metadata required to re-index the perceptron 
entries for training.

In addition to the prefetch table mentioned above, 
PPF also maintains a 1,024-entry direct-mapped ``Reject Table.'' 
If a prefetch suggestion is rejected by the perceptron layer, it is logged into 
the reject table. The table is used to train the perceptron to avoid false negatives 
\textit{i.e.}, cases where the prediction suggested to reject the prefetch 
but the prefetch was proven to be useful based 
on the observed demand accesses to the L2. \newline

\noindent{\textbf{3. Feedback and Data Retrieval}} \newline

When there is an eviction or a demand access to the L2, training for both the base 
prefetcher and our filter mechanism is triggered. 
The address of the cache line that triggered training is used to index
both the Prefetch and Reject tables. 10 bits of the address are used to index 
tables, and another 6 bits to perform tag matching. 
Once the base prefetcher has suggested a prefetch request, the corresponding 
features are used to index the tables of weights. \newline

@@

\noindent{\textbf{4. Training}}\newline

The address from the demand request that triggered the training is lookup in 
both tables. If the address resides in the prefetch table, and is marked 
as valid, this hints the previous prediction was correct and this is a useful
prefetch. We compute the sum of the corresponding weights and
compare the sum to an specific threshold, if the sum falls below it, 
training must happen and the corresponding weights are adjusted accordingly.
These thresholds are introduced in order to avoid over-training, 
which helps the filter adapt quickly to changes in memory behavior. 
These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

On an eviction of a cache block, we look up the corresponding address
in the prefetch table. If there is a valid entry with this address, 
this mean the filter made a misprediction. The block was allocated in the
L2 with a prefetch request that the filter should have categorized 
as a useless prefetch. In this scenario, the corresponding features of the
prefetch request that cause these block to be placed in the cache are used
to index the tables of weights, and those weights are adjusted accordingly.

@@

Parallel to accessing the prefetch table, on a demand access, 
the reject table is accessed. Before
the demand access triggers the next set of prefetches, the reject table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the base prefetcher, but wrongly rejected by the
perceptron filter. The perceptron filter learns from this and makes use of the
corresponding features associated to the original prefetch request, which are
stored in the reject table, to index the weights tables and adjust the weights
accordingly.

Prior work in this domain lacks the knowledge of which cache lines should have 
been prefetched, but were not. The implementation of the reject table,
allows us to have this information, which can be use to further optimize
our prefetching mechanism.

@@

%Section~\ref{Method-Overheads}.

@@

In general, PPF can be adapted to a new base prefetcher with only a
few modifications.
\begin{itemize}

\item \textbf{Enhancing the Base Prefetcher:} By tuning down any
  internal thresholds to increase its inherent aggressiveness.

\item \textbf{Inferencing and Storing:} All prefetch recommendations
  are tested using the perceptron inferencing algorithm. The
  perceptron's output, \textit{true} or \textit{false}, should be
  saved appropriately, along with all metadata required for perceptron
  indexing.

\item \textbf{Retrieving and Training:} When feedback for a prefetch
  is available, the previously stored metadata can be used to re-index
  into the perceptron entries and increment or decrement the weights.

\item \textbf{Feature Selection:}  Perceptrons essentially normalize
  contributions from different features to get a single sum representing the
  final confidence. So the perceptron learning can be as good as the set of 
  features chosen. Interestingly, this is what makes perceptron learning 
  scalable, as it can easily learn to incorporate newer information in 
  the form of new features. Some of the features we developed use information 
  derived directly from program execution, agnostic to the baseline 
  prefetcher. Beyond that, the feature set can be expanded to convey 
  any useful information from the prefetcher to the perceptron filter.

  %[EB] This statement fits better in the next section 'Features used by Perceptron'
  %The methodology explained in
  %Section~\ref{Method-Features} talks about developing a minimal feature 
  %set for PPF.

@@

\end{itemize}

>>>>

\begin{figure}
  \begin{center}
  \includegraphics[width=8cm]{PPF_Hierarchy.pdf}
    \label{fig:PPF_Hierarchy}
  \caption{PPF Architecture in the Memory Hierarchy}
  \end{center}
\end{figure}

It can be beneficial to allow a prefetcher to speculate as deeply as
possible. Often, some useful prefetches are generated long after the
confidence of the prefetcher has fallen below a point, after which the
performance degrades, due to the increase of inaccurate prefetches.  However,
in order to allow deep speculation in the prefetcher, inaccurate prefetches
must be filtered out.  We propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones. The Perceptron Prefetch Filter (PPF) is
placed between the prefetcher and the prefetch insertion queue, as illustrated
in figure ~\ref{fig:PPF_Hierarchy}, to prevent not-useful prefetches from
polluting the higher levels of the memory hierarchy.

Perceptron learning is a light-weight mechanism to pull together disparate
forms of information and synthesize a decision from them.  Our work considers
a number of features corresponding to a prefetch, such as speculation depth,
page address and offset, and uses this information as the inputs to our
perceptron-based filter, in order to predict the usefulness of a prefetch.  In
the remainder of this section, we discuss our proposed technique: PPF.  It
enhances a base prefetcher, by filtering out not-useful prefetches.  PPF is a
generalized prefetch filtering mechanism that may be adapted to any prefetcher
with appropriate modifications.

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as
the steps followed to filter out not-useful prefetches.  The filter is
organized as a set of tables, where each entry in the tables holds a
\textit{weight}. For a configuration of PPF using $N$ number of features, $N$
different tables of weights are needed. Each feature is used to index a
distinct table. The number of entries of each table varies according to their
corresponding feature, hence, different number of bits are needed to index
different tables.  Each weight is a 5-bit saturating counter ranging from -16
to +15.  We found that having 5-bit weights provide a good trade-off between
accuracy and area. A detailed explanation of the storage overhead of PPF can
be found in Figure~\ref{tab:PPF_overhead}.

@@

\noindent{\textbf{Inference}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2 Cache.  At
this point, it has the opportunity to trigger a prefetch. If it decides to do
so, it will also need to decide how many cache blocks to prefetch.  These
blocks can be either placed in the L2 or L3 cache according to the confidence
of the prefetching mechanism.  Once the base prefetcher is triggered, the
suggested prefetch candidates are fed to the perceptron filter to determine
the usefulness of these prefetches.  The filter makes the decision to issue
the prefetch suggestions from the base prefetcher.  To make the decision, each
feature corresponding to a suggested prefetch is used to index a table and all
the corresponding weights are summed.  The sum denotes the confidence value
for the suggested prefetch, and is thresholded against two different
magnitudes: PERC\_THRESHOLD\_HI and PERC\_ THRESHOLD\_ LO.

Prefetches whose sum exceeds PERC\_THRESHOLD\_HI are placed into the L2 cache.
The higher confidence value hints the prefetch would be useful and should be
prioritized.  A prefetch for which the features result in a confidence value
between PERC\_THRESHOLD\_LO and PERC\_THRESHOLD\_HI is allocated in the larger
LLC, as the filter is moderately confident of the future reuse of the cache
block, but not enough to possible pollute a significantly smaller L2.
Suggested prefetches for which the features lead to a confidence value lower
than PERC\_THRESHOLD\_LO are not prefetched, as the low confidence value
represents that the perceptron has learn that similar set of features are
associated with non-useful prefetches. \newline

\noindent{2. Recording}\\
As shown in step 2 of figure ~\ref{fig:PPF_Datapath} the prefetches that make
it through the inference stage are recorded in the ``Prefetch Table''. The
prefetch table is a 1,024-entry, direct mapped structure that contains all
metadata required to re-index the perceptron entries for training.

In addition to the prefetch table mentioned above, PPF also maintains a
1,024-entry direct-mapped ``Reject Table.'' If a prefetch suggestion is
rejected by the perceptron layer, it is logged into the reject table. The
table is used to train the perceptron to avoid false negatives \textit{i.e.},
cases where the prediction suggested to reject the prefetch but the prefetch
was proven to be useful based on the observed demand accesses to the L2. \\

\noindent{\textbf{Feedback and Data Retrieval}\\
When there is an eviction or a demand access to the L2, training for both the
base prefetcher and our filter mechanism is triggered.  The address of the
cache block that triggered training is used to index both the Prefetch and
Reject tables. 10 bits of the address are used to index tables, and another 6
bits to perform tag matching.  Once the base prefetcher has suggested a
prefetch request, the corresponding features are used to index the tables of
weights. \newline

@@

\noindent{\textbf{Training}\newline
The address from the demand request triggering the training is looked up in
both tables. If the address is in the prefetch table and marked as valid, this
hints the previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights.  If the sum falls below a
specific threshold, training occurs and the corresponding weights are adjusted
accordingly.  These thresholds are introduced in order to avoid over-training,
helping the filter adapt quickly to changes in memory behavior.  These
thresholds are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$), respectively for the positive and
negative values of training saturation.

On a cache block eviction, we look up the corresponding address in the
prefetch table. If there is a valid entry with this address, the filter made a
misprediction. The block was allocated in the L2 with a prefetch request that
the filter should have categorized as a useless prefetch.  Thus, the
corresponding features of the prefetch request that cause these block to be
placed in the cache are used to index the tables of weights, and those weights
are adjusted accordingly.

@@

Parallel to accessing the prefetch table, on a demand access, the reject table
is accessed. Before the demand access triggers the next set of prefetches, the
reject table is checked for a valid entry.  A hit means that the corresponding
cache block was initially suggested by the base prefetcher, but wrongly
rejected by the perceptron filter. The perceptron filter learns from this and
makes use of the corresponding features associated to the original prefetch
request, which are stored in the reject table, to index the weights tables and
adjust the weights accordingly.

Prior work in this domain lacks the knowledge of which cache blocks should have 
been prefetched, but were not. The implementation of the reject table,
allows us to have this information, which can be use to further optimize
our prefetching mechanism.

@@

%Section~\ref{Method-Overheads}.  %

@@

In general, PPF can be adapted to a new base prefetcher with only a
few modifications: \\ \\
%
\textbf{Enhancing the Base Prefetcher:} By tuning down any internal thresholds
to increase its inherent aggressiveness. \\
%
\textbf{Inferencing and Storing:} All prefetch recommendations are tested
using the perceptron inferencing algorithm. The perceptron's output,
\textit{true} or \textit{false}, should be saved appropriately, along with all
metadata required for perceptron indexing. \\
%
\textbf{Retrieving and Training:} When feedback for a prefetch is available,
the previously stored metadata can be used to re-index into the perceptron
entries and increment or decrement the weights. \\
%
\textbf{Feature Selection:}  Perceptrons essentially normalize contributions
from different features to get a single sum representing the final confidence.
So the perceptron learning can be as good as the set of features chosen.
Interestingly, this is what makes perceptron learning scalable, as it can
easily learn to incorporate newer information in the form of new features.
Some of the features we developed use information derived directly from
program execution, agnostic to the baseline prefetcher. Beyond that, the
feature set can be expanded to convey any useful information from the
prefetcher to the perceptron filter.

%  %[EB] This statement fits better in the next section 'Features used by
%  Perceptron' %The methodology explained in %Section~\ref{Method-Features}
%  talks about developing a minimal feature %set for PPF.

@@
